Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-18 doc/thesis/methods.tex
More writing

I implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. This algorithm is similar to the reference
\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
the sequence of tolerances $\varepsilon_i$ is automatically determined rather
than specified by the user. The tolerances are chosen such that the \gls{ESS}
of the particle population, which indicates the quality of the Monte Carlo
approximation (see \cref{subsec:sis}), decays at a controlled rate. This
prevents the Monte Carlo approxmiation from collapsing by the variance in
weights growing too large. A single parameter $\alpha$ (not to be confused with
the \gls{BA} model parameter) controls the decay rate, with $\varepsilon_i$
being chosen to satisfy
\[
  \ESS(w_i) = \alpha \ESS(w_{i-1}).
\]
Here, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
depends on $\varepsilon_i$, this equation solves for the updated weights and
the updated tolerance simultaneously. As pointed out by
\textcite{del2012adaptive}, the equation has no analytic solution, but can be
solved by bisection. As in the reference version, the particles are perturbed
by applying an \gls{MCMC} kernel, and resampling occurs when the \gls{ESS}
drops below a prescribed threshold. In my implementation, following all
proposals of the \gls{MCMC} kernel are Gaussian with variance equal to twice
the empirical variance of the
particles~\autocite{beaumont2009adaptive,del2012adaptive}.

The algorithm may be stopped when one of two termination conditions is reached.
The user may specify a final tolerance $\varepsilon$, or a final acceptance
rate of the \gls{MCMC} kernel. The latter stops the algorithm when the
particles are not moving around very much, implying little change in the
estimated target.

>>>>

I implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. This algorithm is similar to the reference
\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
the sequence of tolerances $\varepsilon_i$ is automatically determined rather
than specified in advance. The tolerances are chosen such that the \gls{ESS} of
the particle population, which indicates the quality of the Monte Carlo
approximation (see \cref{subsec:sis}), decays at a controlled rate. This
prevents the Monte Carlo approxmiation from collapsing by the variance in
weights growing too large. A single parameter $\alpha$ (not to be confused with
the \gls{BA} model parameter) controls the decay rate, with $\varepsilon_i$
being chosen to satisfy
\[
  \ESS(w_i) = \alpha \ESS(w_{i-1}).
\]
Here, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
depends on $\varepsilon_i$, this equation solves for the updated weights and
the updated tolerance simultaneously. As pointed out by
\textcite{del2012adaptive}, the equation has no analytic solution, but can be
solved numerically by bisection. The forward kernels $K_i$ are taken to be
\gls{MCMC} kernels with stationary distributions $\pi_{\varepsilon_i}$ and
proposal distributions
\[
  q_i(\theta, \theta') \prod_{k=1}^M \Pr(X_i^{(k)}' \mid \theta'),
\]
where $\theta$ is the vector of model parameters and $X_k$ are $M$ datasets
simulated according to $\theta$. In our implementation, $q$ is a Gaussian
proposal with variance equal to twice the empirical variance of the particles,
following~\autocite{beaumont2009adaptive, del2012adaptive}. The backwards
kernels are
\[
  L_{i-1}(z', z) = \frac{\pi_n(z)K(z, z')}{\pi_n(z')}.
\]
When substituted into \cref{eq:smcwt}, the forward kernels $K(z, z')$ and
densities $\pi_n(z')$ cancel out, and we are left with the weight update
\begin{align*}
  w(z') 
    &\propto w(z) \frac{\pi_n(z \mid y)}{\pi_{n-1}(z \mid y)} \\
    &= w(z) \frac{\pi(z) \pi_n(y \mid z)}{\pi(z) \pi_{n-1}(y \mid z)} \\
    &= w(z) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_n, y}}(x_k)}
            {\sum_{k=i}^M \I_{A_{\varepsilon_{n-1}, y}}(x_k)}.
\end{align*}
The algorithm may be stopped when one of two termination conditions is reached.
The user may specify a final tolerance $\varepsilon$, or a final acceptance
rate of the \gls{MCMC} kernel. The latter stops the algorithm when the
particles are not moving around very much, implying little change in the
estimated target.
